Reverse eigenvalue problem

### 2024-03-26 Docs/NhanesApplExample.Rmd
repeated simulation use NHANES estimates

```{r fGFPCA}
# Step 1: Bin every 10 observations
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

@@

# mean 
# because of the parametrization of mgcv
# we need to add the constant intercept back
true_mu <- predict(debias_glmm, type = "terms")[, 1]+coef(debias_glmm)[1]
true_mu <- matrix(true_mu, nrow = J)[, 1]
# plot(1:J, true_mu)

@@

## Data generation

```{r gen_data_unbound}
# interpolate FPCA mean
# interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
# true_mu <- interp_mu_fpca$y

# increase sample size?
Ntry <- 500
# generate score
xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))

@@

# binary outcome
gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
```

```{r plot_data}
# comparing generated data to the actual data
bind_rows(
    gen_df %>% 
      group_by(sind) %>% 
      summarize(active = mean(Y)) %>%
      mutate(type="Simulation"),

    df %>% 
      group_by(sind) %>% 
      summarize(active = mean(Y)) %>%
      mutate(type="NHANES")
) %>%
  ggplot()+
      geom_line(aes(x=sind, y=active, col = type))+
      labs(title="Proportion of active subjects by minute")

@@

## fGFPCA

```{r exp_gen_data}
# bin data
# using the same bin length 
gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
gen_df$bin <- as.numeric(as.character(gen_df$bin))

# Step 2:  Local GLMM
gen_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)
## Figure
df_est_latent %>% filter(id %in% 1:4) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
  geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
  facet_wrap(~id)

# FPCA
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                        nrow=Ntry, 
                        ncol=n_bin, byrow = F) 
# dim(mat_est_unique)
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
```

```{r}
# Re-evaluation
est_phi <- matrix(NA, J, K) 
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

@@

### model fit with B-splines basis
t1 <- Sys.time()
this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
t2 <- Sys.time()
```

Below I compared the estimated eigenvalues from simulated data to the true eigenvalues. As we see, the reverse did not happen on the simulated data.

```{r new_evals, class.source='fold-show'}
# estimated eigenvalues
1/this_glmm$sp[2:5]/n_bin

# true eigenvalues
true_evals
```
Below is the comparision between true and esimtated eigenfunctions.

```{r new_PC}
bind_rows(df_phi %>% mutate(type = "True"),
          est_phi %>% mutate(type = "Estimate",
                             phi1 = phi1*sqrt(n_bin),
                             phi2 = phi2*sqrt(n_bin),
                             phi3 = phi3*sqrt(n_bin),
                             phi4 = phi4*sqrt(n_bin))) %>%
  pivot_longer(2:5) %>%
  ggplot()+
  geom_line(aes(x=sind, y=value, col = type))+
  facet_wrap(~name)
```

And below is the comparision between true mean, mean estimated from FPCA, and re-evaluated mean from the BAM model in Step 4. The re-evaluation seem to drive the mean function closer to the truth. But in that case, why would generated data much more active compared to the NHANES data?

```{r compare_mean}
bind_rows(
  data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"),
  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]+coef(this_glmm)[1]) %>% mutate(type = "BAM"),
  data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA")
) %>%
  ggplot()+
      geom_line(aes(x=sind, y=mu, col = type))+
      labs(title="Mean estimates")
```

>>>>

```{r}
# Step 1: Bin every 10 observations
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

@@

# mean 
# because of the parametrization of mgcv
# we need to add the constant intercept back
true_mu <- predict(debias_glmm, type = "terms")[1:J, 1]+coef(debias_glmm)[1]
# plot(1:J, true_mu)

@@

## Simulation

Let's repeat the simulation a few times:

- With different sample sizes
- With a lower-bounded mean

### Sample size

```{r gen_data_N}
# increase sample size
Ntry_vec <- c(500, 1000, 1500, 2000)
data_list <- list()

for(n in seq_along(Ntry_vec)){

  Ntry <- Ntry_vec[n]

  # generate score
  xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))

@@

  # binary outcome
  gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))

  data_list[[n]] <- gen_df

}
```

```{r plot_data, message=FALSE}
data_list %>% bind_rows(.id = "size") %>%
  mutate(size = factor(size, levels = 1:length(Ntry_vec), labels = Ntry_vec)) %>% 
  group_by(size, sind) %>% 
  summarize(active = mean(Y)) %>%
  left_join(df %>% group_by(sind) %>% summarize(true_active = mean(Y)),
            by = "sind") %>%
  ggplot()+
  geom_line(aes(x=sind, y=active, col = size), alpha = 0.5)+
  geom_smooth(aes(x=sind, y=active, col = size), linewidth = 0.5)+
  geom_line(aes(x=sind, y=true_active), col = "black", 
            linewidth=0.5)+
  labs(title="Proportion of active subjects by minute")

@@

```{r}
# containers
time_vec <- rep(NA, length(Ntry_vec))
fpca_evals <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
bam_evals <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
```

```{r fGFPCA}
for(n in seq_along(Ntry_vec)){

  Ntry <- Ntry_vec[n]
  gen_df <- data_list[[n]]

  # bin data
  # using the same bin length 
  gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
  gen_df$bin <- as.numeric(as.character(gen_df$bin))

  # Step 2:  Local GLMM
  gen_bin_lst <- split(gen_df, f = gen_df$bin)
  df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
  df_est_latent <- bind_rows(df_est_latent)

  # FPCA
  uni_eta_hat <- df_est_latent %>% filter(bin==sind)
  mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                          nrow=Ntry, 
                          ncol=n_bin, byrow = F) 
  fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs

  # Re-evaluation
  est_phi <- matrix(NA, J, K) 
  p <- 3 # order of b splines 
  knots <- 35 # number of knots (same from FPCA model)
  knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
  knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

@@

  ### model fit with B-splines basis
  t1 <- Sys.time()
  this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
  t2 <- Sys.time()

  time_vec[n] <- t2-t1
  fpca_evals[n, ] <- fpca_mod$evalues
  bam_evals[n, ] <- 1/this_glmm$sp[2:5]/n_bin

}
```

- eigenvalues from FPCA

```{r}
# time_vec
data.frame(size = Ntry_vec, fpca_evals)
```

- eigenvalues from BAM

```{r}
data.frame(size = Ntry_vec, bam_evals)
```

- true eigenvalues used to generate data

```{r}
# true eigenvalues
true_evals
```

<!-- Below is the comparision between true and esimtated eigenfunctions.  -->

<!-- ```{r new_PC} -->
<!-- bind_rows(df_phi %>% mutate(type = "True"), -->
<!--           est_phi %>% mutate(type = "Estimate", -->
<!--                              phi1 = phi1*sqrt(n_bin), -->
<!--                              phi2 = phi2*sqrt(n_bin), -->
<!--                              phi3 = phi3*sqrt(n_bin), -->
<!--                              phi4 = phi4*sqrt(n_bin))) %>% -->
<!--   pivot_longer(2:5) %>% -->
<!--   ggplot()+ -->
<!--   geom_line(aes(x=sind, y=value, col = type))+ -->
<!--   facet_wrap(~name) -->
<!-- ``` -->

<!-- And below is the comparision between true mean, mean estimated from FPCA, and re-evaluated mean from the BAM model in Step 4. The re-evaluation seem to drive the mean function closer to the truth. But in that case, why would generated data much more active compared to the NHANES data?  -->

<!-- ```{r compare_mean} -->
<!-- bind_rows( -->
<!--   data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"), -->
<!--   data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]+coef(this_glmm)[1]) %>% mutate(type = "BAM"), -->
<!--   data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA") -->
<!-- ) %>% -->
<!--   ggplot()+ -->
<!--       geom_line(aes(x=sind, y=mu, col = type))+ -->
<!--       labs(title="Mean estimates") -->
<!-- ``` -->

## Bounded mean

- Here, we put a lower bound to the mean function, so that the probablity of active status is greater than 0.2

```{r}
bd_mean <- ifelse(plogis(true_mu)>=0.2, true_mu, logit(0.2))
plot(1:J, plogis(true_mu), type = "l", col = "black", 
     main = "Population mean on the probablity scale",
     xlab = "sind", ylab = "")
lines(1:J, plogis(bd_mean), col = "red")

```
```{r gen_data_bd_mean}
# increase sample size
data_list_bdmean <- list()

for(n in seq_along(Ntry_vec)){

  Ntry <- Ntry_vec[n]

  # generate score
  xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))

  # add in unbounded mean
  eta <- apply(bi, 2, function(x){x+bd_mean})
  gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
                       sind = rep(1:J, Ntry),
                       eta_i = as.vector(eta))

  # binary outcome
  gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))

  data_list_bdmean[[n]] <- gen_df

}
```

```{r plot_data_bdmean, message=FALSE}
data_list_bdmean %>% bind_rows(.id = "size") %>%
  mutate(size = factor(size, levels = 1:length(Ntry_vec), labels = Ntry_vec)) %>% 
  group_by(size, sind) %>% 
  summarize(active = mean(Y)) %>%
  left_join(df %>% group_by(sind) %>% summarize(true_active = mean(Y)),
            by = "sind") %>%
  ggplot()+
  geom_line(aes(x=sind, y=active, col = size), alpha = 0.5)+
  geom_smooth(aes(x=sind, y=active, col = size), linewidth = 0.5)+
  geom_line(aes(x=sind, y=true_active), col = "black", 
            linewidth=0.5)+
  labs(title="Proportion of active subjects by minute")

```{r}
# containers
fpca_evals_bdmean <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
bam_evals_bdmean <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
```

```{r fGFPCA_bdmean}
for(n in seq_along(Ntry_vec)){

  Ntry <- Ntry_vec[n]
  gen_df <- data_list_bdmean[[n]]

  # bin data
  # using the same bin length 
  gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
  gen_df$bin <- as.numeric(as.character(gen_df$bin))

  # Step 2:  Local GLMM
  gen_bin_lst <- split(gen_df, f = gen_df$bin)
  df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
  df_est_latent <- bind_rows(df_est_latent)

  # FPCA
  uni_eta_hat <- df_est_latent %>% filter(bin==sind)
  mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                          nrow=Ntry, 
                          ncol=n_bin, byrow = F) 
  fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs

  # Re-evaluation
  est_phi <- matrix(NA, J, K) 
  p <- 3 # order of b splines 
  knots <- 35 # number of knots (same from FPCA model)
  knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
  knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

  ### model fit with B-splines basis
  t1 <- Sys.time()
  this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
  t2 <- Sys.time()

  # time_vec[n] <- t2-t1
  fpca_evals_bdmean[n, ] <- fpca_mod$evalues
  bam_evals_bdmean[n, ] <- 1/this_glmm$sp[2:5]/n_bin

}
```

- eigenvalues from FPCA

```{r}
# time_vec
data.frame(size = Ntry_vec, fpca_evals_bdmean)
```

- eigenvalues from BAM

```{r}
data.frame(size = Ntry_vec, bam_evals_bdmean)
```
